The Evolution of Points Scored in the NBA

### 2024-04-19 paper/paper.qmd
final push

The evolution of scoring in basketball throughout its tenure has brought about various discussions between the older generation of fans and the newer generation. A main topic of debate is in regards to the scoring of the game. Throughout this paper, we will be looking into the statistical data from the 2000-2001 NBA season up to the current 2023-2024 season. By analyzing the data and considering the changes, we aim to deduce the significant factors that have impacted scoring in the NBA. There are also other factors that may not be as obvious in the statistics, such as the change in pace of the game, how defenses have changed, and how the playstyles have evolved. This shows that what most would have as the 'conclusive main argument' for the 'rise of NBA scoring' in three-point shooting, may not be as significant on its own as it was originally perceived. There are many more factors that come into play that may co-align with the rise of three-point shooting, but have not been properly considered. All of these understandings are vital in predicting the possible future of NBA scoring. Suppose an NBA team replaced every single NBA shot with a three-point shot, since it is worth more points, then the entire game plan would change. Opposing teams would not consider that team a threat when they are closer to the basket, so they could overload the three point line and work to lower their opponents' three-point shooting percentage beyond the league average. This hypothetical scenario could end up with that three-point shooting team scoring significantly less than the average NBA team. 'Live by the three, die by the tree', as they would call it.

The estimand is what were investigating as we look for the main factors, both offensive and defensive that contribute to the rise in NBA points scored.

These are all factors that we will be looking into throughout this paper. This paper aims to make use of @citeR to provide an analysis of the main factors that have contributed to the rise in scoring in the NBA and how it may affect the future of the sport, from statistics, to the player archetypes sought after, to whether or not the current NBA records will be broken. After introducing the chosen statistical data ranging from individual teams and players, to cumulative, league-wide stats, we will incorporate this data into a model to predict whether or not scoring will continue to rise, or if it has reached its peak.

@@

## First discussion point {#sec-first-point}

## Second discussion point {#sec-sec-point}

## Third discussion point {#sec-third-point}

## Weaknesses and next steps

>>>>

The evolution of scoring in basketball throughout its tenure has brought about various discussions between the older generation of fans and the newer generation. A main topic of debate is in regards to the scoring of the game. Throughout this paper, we will be looking into the statistical data from the 2000-2001 NBA season up to the current 2023-2024 season. By analyzing the data and considering the changes, we aim to deduce the significant factors that have impacted scoring in the NBA. There are also other factors that may not be as obvious in the statistics, such as the change in pace of the game, how defenses have changed, and how the playstyles have evolved. This shows that what most would have as the 'conclusive main argument' for the 'rise of NBA scoring' in three-point shooting, may not be as significant on its own as it was originally perceived. There are many more factors that come into play that may co-align with the rise of three-point shooting, but have not been properly considered. All of these understandings are vital in predicting the possible future of NBA scoring. Suppose an NBA team replaced every single NBA shot with a three-point shot, since it is worth more points, then the entire game plan would change. Opposing teams would not consider that team a threat when they are closer to the basket, so they could overload the three point line and work to lower their opponents' three-point shooting percentage beyond the league average. This hypothetical scenario could end up with that three-point shooting team scoring significantly less than the average NBA team. 'Live by the three, die by the tree', as they would call it. The problem is, we don't directly know what is the main source behind this surge in scoring, or if it is a single factor alone.

Shooting the basketball, affecting the offense through intangibles, and defensive effort, these are all factors that we will be looking into throughout this paper. This paper aims to make use of @citeR to provide an analysis of the main factors that have contributed to the rise in scoring in the NBA and how it may affect the future of the sport, from statistics, to the player archetypes sought after, to whether or not the current NBA records will be broken. After introducing the chosen statistical data ranging from individual teams and players, to cumulative, league-wide stats, we will incorporate this data into a generalized linear model to predict whether or not scoring will continue to rise, or if it has reached its peak. After the creation of four models, we can highlight any significant instances that make a big impact towards scoring. Whether it is making a play on the defensive end to score a fast-break bucket, or shooting the ball at a higher percentage, the main purpose of the paper is to break it down into simple, basic NBA metrics to get a better understanding of how the game has changed. The estimand is what were investigating as we look for the main factors, both offensive and defensive that contribute to the rise in NBA points scored.

@@

## First discussion point {#sec-first-point}
The first discussion point is how significant three point shooting is to NBA offense. It was a clear factor to keep in mind prior to developing the paper. Many people hear about how Stephen Curry has changed the game forever as the NBA offense transitioned to shooting more threes. This was evident on Figure 2 in the Data section as they showed the clear correlation between three pointers attempted and points scored, as well as an emphasis that it may not just be better shooting, but rather the volume of threes. This gives the idea that if you can shoot more threes, while you do expect to miss more shots than a more efficient mid-range, you expect more points throughout the entire game. If you shoot 10 threes at 35% that is approximately 10.5 points, while if you shoot 10 field goals at 45%, that is approximately 9 points scored. This shows how NBA teams have adapted to accomodate for more statistically efficient methods of shooting.

## Second discussion point {#sec-sec-point}
The second discussion point is how the offensive intangibles have affected NBA offense. It is clear from Figure 4 from the Data section since it showed the correlation between assists and points scored. There are certainly more ways than scoring to affect an NBA offense. There are a ton of ways that players can impact an opposing team's gameplan. If a player is a good shooter, then the opposing defender will stick closer to them allowing more space in the area closer to the hoop. This is called spacing the floor, an intangible offensive asset. Increased ball movement can also lead to better offensive plays because it can allow you to get a better look (more wide open from defender's contest) for a teammate which can be a higher percentage shot. There are a lot of ways to impact an offense, and offensive intangibles are one of them, but it is clear that they are not as significant as shooting the ball itself.

## Third discussion point {#sec-third-point}
The last point of discussion is regarding defense. What I failed to consider when creating my model was that the defense of one NBA team impacts the offense of the other NBA team. This is a tradeoff that is reflected on the overall team stats for the season. This is why I was shocked to see those values in the defensive model summary, but I understood that this was an oversight on my end. There may be another way to measure the change in defense over the years and correlate it to the increase in offensive output. Possibly look towards seeing if there is a lack of defense being played in the league that is not reflected in numbers alone? Watching games can help with this as you can see whether players are being left wide open in games. There is also the point to consider that handchecking was banned in the NBA in the mid 2000s, which could have possibly made a great impact on how NBA offense is played since defenses are not allowed to be as physical.

## Weaknesses and next steps
There is only so much you can portray when you have statistics only. There are a lot of parts of basketball that cannot be measured in stats alone. One big thing to consider now is that basketball is more popular than ever. There are a lot of players that begin young and are more prepared for when they play in the NBA. You could consider this an increase in talent as well. While this does not guarantee that you will see an anomaly such as the next Michael Jordan, or LeBron James, it does increase the floor of NBA players and thus, increases the overall talent in the league. This could be why there is an increase in offensive output, but it is not a factor that we can measure.
